Investigating shape standardisation using geometric morphometry of pottery from Iron Age northeastern Taiwan

### 2020-04-10 analysis/paper/paper.Rmd
edits to gmm section

Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interest. Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as xy coordinates for two dimensions, or xyz coordinates for three dimensions. In a typical analysis, starting with raw landmarks, a sample of objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistical analyses [@Slice2007]. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or biological measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contain crucial variation. Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by specialists attached to elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottles excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods. Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on Indigenous pottery production that can be detected in the shape of the vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?

>>>>

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes (Adams et al., 2004; Bookstein, 1997; Lawing and Polly, 2010; Slice, 2007). Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark approaches and outline approaches (Adams et al., 2004). Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features (Birch and Martinón-Torres, 2019; Buchanan et al., 2019; Cardillo, 2010; Haruda et al., 2019; Lycett and Cramon-Taubadel, 2013; Meloro et al., 2015).

However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals (Bookstein, 1997; Lawing and Polly, 2010). Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy (Bookstein, 1997; Gunz and Mitteroecker, 2013; Slice, 2007). Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients (Kuhl and Giardina, 1982). EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions (Adams et al., 2004; Bonhomme et al., 2014; Claude, 2008). The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics (Bonhomme et al., 2014). EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics (Fox, 2015; Hoggard et al., 2019; Ioviţă, 2010; Topi et al. 2018; Selden Jr 2019; Wilczek et al. 2014). Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
